Circumnuclear Media of Quiescent Supermassive Black Holes

### 2015-06-29 ms.tex
not to self

We fix $\beta = 2$ but find that our results are not overly sensitive
to the properties of the gas flow on radial scales $\gtrsim r_b$ for
values of $\beta\gsim 1.5$, which characterize most of the
\citet{LauerFaber+:2007a} sample\footnote{The median $\beta$ in the
  \citet{LauerFaber+:2007a} sample is 1.39 (1.64) for core (cusp)
  galaxies}.  The presence of the break radius $r_{b}$ is, however,
necessary to obtain a converged steady state for some regions of our
parameter space.  We consider solutions calculated for up to four
values of $\rb$: 50 pc, 100 pc, 200 pc, and 400 pc, motivated by the
range of break radii from the \citet{LauerFaber+:2007a}
sample.\footnote{For the core galaxies the break radius follows the
  scaling relationship $\rb\sim 106 \, (\Mbheight)^{0.39}$ pc, with
  scatter of approximately one dex.  Most the cusp galaxies have $\rb$
  between 100 and 1000 pc, and lack a clear trend with $\Mbh$.}  We
neglect values of $r_b$ which would give unphysically large bulges for
a given $\Mbh$.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{ti.pdf}
  \caption{Minimum effective wind heating parameter required for
    thermal stability as a function of SMBH mass.  Black lines show
    $v_{\rm TI}$ (eq.~[\ref{eq:cooling3}]), the heating rate required
    for $(\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|)_{r_{s}} > 10$ in the
    high-heating limit when the stagnation radius lies interior to the
    influence radius, for different values of the mass loss parameter
    $\eta$ as marked.  Blue lines show the minimum heating parameter
    required to have $\zeta > \zeta_c = (r_{b}/r_{\rm
      inf})^{0.5(1-\Gamma)}$ (eq.~[\ref{eq:zetacrit}]), separately for
    cusp (solid) and core (dashed) galaxies. Based on the
    \citet{LauerFaber+:2007a} sample we take $\rinf=25 (8) \Mbh^{0.6}
    pc$ and $\rb=106 \Mbheight^{0.4}$ pc (100 pc)  for cores (cusps). For
    $\zeta < \zeta_c$ the stagnation radius moves from inside the
    influence radius, out to the stellar break radius $r_b$ (Appendix
    \ref{app:be}); in practice renders the flow susceptible to thermal
    runaway, even if $v_{w} > v_{\rm TI}$.  For cusp galaxies the $v_w
    > v_{\rm TI}$ is usually the more strigent criterion for
    stability, while for core galaxies the $\zeta > \zeta_c$ is
    usually more relevant.}
\label{fig:TI}
\end{figure}

@@

\citet{Sharma+2007} estimate the value of the radiative efficiency of
low luminosity AGN, $\epsilon_{\rm rad}$, based on MHD shearing box
simulations of collisionless plasmas.  For Eddington ratios of
relevance, their results (shown in their Fig.~6) are well approximated
by
\begin{align}
\epsilon_{\rm rad} \simeq 
\begin{cases}
  0.03 \left(\frac{\dot{M}_{\bullet}}{10^{-4}\dot{M}_{\rm edd}}\right)^{0.9} & \frac{\dot{M}_{\bullet}}{\dot{M}_{\rm edd}} \lsim 10^{-4} \\
 0.03 &  10^{-2} \lsim \frac{\dot{M}_{\bullet}}{\dot{M}_{\rm edd}} \lsim  10^{-4},
\end{cases}
\label{eq:efficiency}
\end{align}
where $\dot{M}_{\bullet}$ is the BH accretion rate. In general
$\dot{M}_{\bullet}$ will be smaller than the inflow rate $\dot{M}$
calculated thus far by a factor $\alpha < 1$ due to outflows from the
accretion disk on small scales.  Thus, the full efficiency relating
the mass inflow rate to the radiative output is $\epsilon = \alpha
\epsilon_{\rm rad}$, where we take $\alpha = 0.1$ following
\citet{Li+2013}, who find that the fraction of the inflowing matter
lost to outflows equals the Shakura-Sunyaev viscosity parameter of the
disk.

Substituting $\epsilon$ into equation (\ref{eq:MdotCa}) and solving
for the inflow rate results in
\begin{align}
\frac{\dot{M}_{C}}{\dot{M}_{\rm edd}} \approx 
\begin{cases} 6\times 10^{-3} \eta_{0.02}^{0.12}T_{\rm
C,9}^{-0.44}M_{\bullet,8}^{0.09}
&, \text{core}\\ 2\times 10^{-3} \eta_{0.02}^{0.17}T_{\rm
C,9}^{-0.4} M_{\bullet,8}^{0.08}
&, \text{cusp}.
  \end{cases}
  \label{eq:MdotC}
\end{align}
Accretion may not be truly steady in cases where AGN heating
dominates, due to the inherent delay between black hole feedback and
the structure of the accretion flow on larger scales.  Equation
(\ref{eq:MdotC}) may nevertheless represent a characteristic average
value for the inflow rate if a quasi-equilibrium is achieved over
many cycles.

@@

Strikingly, we find that galactic nuclei that experience the same
average SFH as their host galaxies possess thermally stable flows
across all black hole masses.  An important caveat, however, is that
$\vw$ is generally only a few times higher than the stellar velocity
dispersion, i.e. $\zeta \sim \zeta_c$ (eq.~[\ref{eq:zetacrit}]).  This
implies that the true stagnation radius and the inflow rate could
be larger than our analytic estimates, potentially resulting in
thermal instability in a significant fraction of galaxies.  The bottom
panel of Figure~\ref{fig:vwSources} also shows $\zeta/\zeta_c$, where
$\zeta_c$ is estimated from the best-fit power-law dependence of
$r_b$, $\rinf$, and $\Gamma$ on $M_{\bullet}$ from the
\citet{LauerFaber+:2007a} galaxy sample.  In other words, the CNM of a
significant fraction of massive ellipticals may be thermally unstable,
not necessarily because they are receiving lower stellar feedback than
the average for their galaxy mass, but due to the high heating
required for stability given the structure of their stellar potential.

@@

The unresolved core X-ray luminosities of nearby galactic nuclei
provide a powerful diagnostic of the SMBH accretion rates and how they
vary with SMBH mass and other galaxy properties (e.g., \citealt{Ho08}
and references therein).  Figure~\ref{fig:miller} shows our
predictions for $\langle L_{X} \rangle =\epsilon_X \dot{M} c^2$ as a
function of black hole mass, where $\epsilon_{X} = \alpha\epsilon_{\rm
  rad}\epsilon_{\rm bol}$, $\alpha = 0.1$ accounts for the fraction of
the inflowing matter loss to outflows from the accretion disk,
$\epsilon_{\rm rad}$ is the radiative efficiency of the accreted
matter, and $\epsilon_{\rm bol} = 0.1$ is the assumed bolometric
correction into the measured X-ray band for low luminosity AGN
(\citealt{Ho08}).  In all cases the value of $\langle L_X \rangle$ is
calculated using the accretion rate from the top line of
equation~\eqref{eq:mdot_analytic} with $\Gamma = 0.3
\Mbheight^{-0.24}$, based on fitting the \citet{LauerFaber+:2007a}
galaxy sample.

@@

\begin{figure*}
\includegraphics[width=\textwidth]{miller.pdf}
\caption{\label{fig:miller} Average nuclear X-ray luminosity, $\langle
  L_X \rangle = \epsilon_X \Mdot c^2$, as a function of SMBH mass.
  Green lines show our prediction in the case of heating due to the
  average SFH for galaxies corresponding to each black hole mass
  (Fig.~\ref{fig:bh_growth}), calculated using a power-law fit for the
  relationship $\Gamma(M_{\bullet})=0.3 \Mbheight^{-0.24}$ derived
  from the \citet{LauerFaber+:2007a} sample.  Shown for comparison are
  the measurements ({\it black stars}) and upper limits ({\it gray
    triangles}) for $L_X/L_{\rm edd}$ values, from the
  \citet{Miller+15} sample of early-type galaxies (a black line shows
  the best power-law fit $\langle L_X/L_{\rm edd}\rangle \propto
  M_{\bullet}^{\alpha}$, with $\alpha = -0.2$).  The left panel is
  calculated assuming a constant radiative efficiency
  $\epsilon_X=10^{-4} (\epsilon_{\rm bol}/0.1)$, while the right hand
  panel assumes $\epsilon_X= \alpha \epsilon_{\rm bol}\epsilon_{\rm
    rad}$, where $\epsilon_{\rm rad}$ is the radiative efficiency of
  low luminosity accretion disks calculated by \citet{Sharma+2007}
  (eq.~[\ref{eq:efficiency}]) and $\alpha = 0.1$ is the fraction of
  the mass inflowing on large scales that reaches the SMBH.  Shown for
  comparison are the X-ray luminosities calculated for the maximum
  thermally-stable accretion rate ({\it dashed orange line};
  eq.~[\ref{eq:Mdotmax}]); the SN Ia-regulated accretion rate ({\it
    dashed teal line}; eq.~[\ref{eq:eddr_Ia}]), and the Compton
  heating-regulated accretion rate ({\it dashed pink line};
  eq.~[\ref{eq:MdotCa}]), again all calculated for the average SFH
  corresponding to each SMBH mass.}
\end{figure*}

If the nuclei of elliptical galaxies are heated as expected for the
average SFH of galaxies with similar mass, then to first order we
predict that $\langle L_{X}/L_{\rm edd}\rangle$ should be a weakly
increasing function of BH mass.  This result is in tension with the
observed down-sizing trend that the average Eddington ratio is seen to
decrease (albeit weakly) with $M_{\bullet}$, especially in the model where
$\epsilon_X$ increases with the Eddington ratio.
However, one must keep in mind the enormous uncertainty in calculating
the luminosity of the accretion flow close to the black hole in a
single waveband based on the feeding rate on larger scales.  Also, if disk
outflows indeed carry away most of the infalling mass before it reaches X-ray
producing radii (e.g.~\citealt{Blandford&Begelman99};
\citealt{Li+13}), then the angular momentum of the infalling gas might
also influence the X-ray luminosity indirectly through the inflow efficiency $\alpha$, in a way that could depend systematically on the stellar population and hence $M_{\bullet}$.  However, in general the efficiency with which inflowing matter reaches the SMBH (and hence X-ray luminosity) would naively be expected to {\it decrease} with decreasing $M_{\bullet}$ due to the increasing angular momentum of low mass galaxies, exacerbating the tension between our findings and the observed downsizing trend.

@@

\item Stellar wind heating from the average SFH may be sufficient to
  permit thermally-stable, steady-state accretion, depending on the
  accretion efficiency of the SMBH.  However, the resulting average
  Eddington-normalized accretion rate is predicted to increase weakly
  with $M_{\bullet}$ (Fig.~\ref{fig:miller}), in tension with the
  (also weak) downsizing trend of measured X-ray luminosities in
  early-type galactic nuclei (e.g., \citealt{Miller+15}).  Thermally
  stable accretion models can reproduce the observed scatter in
  nuclear X-ray luminosities at fixed $\Mbh$ (2-4 orders of magnitude)
  due to the combination of (1) differences in the stellar wind
  heating rate due to stochastic variation in SFH between galaxies, or
  within one galaxy due to burstiness in the SFH, as in
  ~Fig.~\ref{fig:NickPlot2}; (2) Variations in the amount of inflowing
  mass which is lost to small scale outflows from the SMBH accretion
  disk, likely due to variations in the angular momentum of the
  accreting gas (cf.~\citealt{Pellegrini10}).

>>>>

We fix $\beta = 2$ but find that our results are not overly sensitive
to the properties of the gas flow on radial scales $\gtrsim r_b$ for
values of $\beta\gsim 1.5$, which characterize most of the
\citet{LauerFaber+:2007a} sample\footnote{The median $\beta$ in the
  \citet{LauerFaber+:2007a} sample is 1.39 (1.64) for core (cusp)
  galaxies}.  The presence of the break radius $r_{b}$ is, however,
necessary to obtain a converged steady state for some regions of our
parameter space.  We consider solutions calculated for up to four
values of $\rb$: 50 pc, 100 pc, 200 pc, and 400 pc, motivated by the
range of break radii from the \citet{LauerFaber+:2007a}
sample.\footnote{For the core galaxies the break radius follows the
  scaling relationship $\rb\sim 106 \, (\Mbheight)^{0.39}$ pc, with
  scatter of approximately one dex.  Most the cusp galaxies have $\rb$
  between 100 and 1000 pc, and lack a clear trend with $\Mbh$. {\bf AG
    I redid the fitting and found $\rb\sim 90 \, (\Mbheight)^{0.5}$
    pc}} We neglect values of $r_b$ which would give unphysically
large bulges for a given $\Mbh$.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{ti.pdf}
  \caption{Minimum effective wind heating parameter required for
    thermal stability as a function of SMBH mass.  Black lines show
    $v_{\rm TI}$ (eq.~[\ref{eq:cooling3}]), the heating rate required
    for $(\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|)_{r_{s}} > 10$ in the
    high-heating limit when the stagnation radius lies interior to the
    influence radius, for different values of the mass loss parameter
    $\eta$ as marked.  Blue lines show the minimum heating parameter
    required to have $\zeta > \zeta_c = (r_{b}/r_{\rm
      inf})^{0.5(1-\Gamma)}$ (eq.~[\ref{eq:zetacrit}]), separately for
    cusp (solid) and core (dashed) galaxies. Based on the
    \citet{LauerFaber+:2007a} sample we take $\rinf=25 (8) \Mbh^{0.6}
    pc$ and $\rb=106 \Mbheight^{0.4}$ pc (100 pc)  for cores (cusps)
    {\bf AG update rb fit}. For
    $\zeta < \zeta_c$ the stagnation radius moves from inside the
    influence radius, out to the stellar break radius $r_b$ (Appendix
    \ref{app:be}); in practice renders the flow susceptible to thermal
    runaway, even if $v_{w} > v_{\rm TI}$.  For cusp galaxies the $v_w
    > v_{\rm TI}$ is usually the more strigent criterion for
    stability, while for core galaxies the $\zeta > \zeta_c$ is
    usually more relevant.}
\label{fig:TI}
\end{figure}

@@

\citet{Sharma+2007} estimate the value of the radiative efficiency of
low luminosity AGN, $\epsilon_{\rm rad}$, based on MHD shearing box
simulations of collisionless plasmas.  For Eddington ratios of
relevance, their results (shown in their Fig.~6) are well approximated
by
\begin{align}
\epsilon_{\rm rad} \simeq 
\begin{cases}
  0.03 \left(\frac{\dot{M}_{\bullet}}{10^{-4}\dot{M}_{\rm edd}}\right)^{0.9} & \frac{\dot{M}_{\bullet}}{\dot{M}_{\rm edd}} \lsim 10^{-4} \\
 0.03 &  10^{-2} \gsim \frac{\dot{M}_{\bullet}}{\dot{M}_{\rm edd}} \gsim  10^{-4},
\end{cases}
\label{eq:efficiency}
\end{align}
where $\dot{M}_{\bullet}$ is the BH accretion rate. In general
$\dot{M}_{\bullet}$ will be smaller than the inflow rate $\dot{M}$
calculated thus far by a factor $\alpha < 1$ due to outflows from the
accretion disk on small scales.  Thus, the full efficiency relating
the mass inflow rate to the radiative output is $\epsilon = \alpha
\epsilon_{\rm rad}$, where we take $\alpha = 0.1$ following
\citet{Li+2013}, who find that the fraction of the inflowing matter
lost to outflows equals the Shakura-Sunyaev viscosity parameter of the
disk.

Substituting $\epsilon$ into equation (\ref{eq:MdotCa}) and solving
for the inflow rate results in

\begin{align}
\frac{\dot{M}_{C}}{\dot{M}_{\rm edd}} \approx 
\begin{cases} 4\times 10^{-2} \eta_{0.02}^{0.2}T_{\rm
C,9}^{-0.8} M_{\bullet,8}^{0.16}
&, \text{core}\\ 9\times 10^{-3} \eta_{0.02}^{0.3}T_{\rm
C,9}^{-0.7} M_{\bullet,8}^{0.14}
&, \text{cusp}.
  \end{cases}
  \label{eq:MdotC}
\end{align}

Accretion may not be truly steady in cases where AGN heating
dominates, due to the inherent delay between black hole feedback and
the structure of the accretion flow on larger scales.  Equation
(\ref{eq:MdotC}) may nevertheless represent a characteristic average
value for the inflow rate if a quasi-equilibrium is achieved over
many cycles.

@@

Strikingly, we find that galactic nuclei that experience the same
average SFH as their host galaxies possess thermally stable flows
across all black hole masses.  An important caveat, however, is that
$\vw$ is generally only a few times higher than the stellar velocity
dispersion, i.e. $\zeta \sim \zeta_c$ (eq.~[\ref{eq:zetacrit}]).  This
implies that the true stagnation radius and the inflow rate could
be larger than our analytic estimates, potentially resulting in
thermal instability in a significant fraction of galaxies.  The bottom
panel of Figure~\ref{fig:vwSources} also shows $\zeta/\zeta_c$, where
$\zeta_c$ is estimated ...{\bf AG this sentence needs to be
  fixed. Power law fits different for cores and cusps. Power law fixed
to be 0.6 for influence radius in doing fits}.  In other words, the CNM of a
significant fraction of massive ellipticals may be thermally unstable,
not necessarily because they are receiving lower stellar feedback than
the average for their galaxy mass, but due to the high heating
required for stability given the structure of their stellar potential.

@@

The unresolved core X-ray luminosities of nearby galactic nuclei
provide a powerful diagnostic of the SMBH accretion rates and how they
vary with SMBH mass and other galaxy properties (e.g., \citealt{Ho08}
and references therein).  Figure~\ref{fig:miller} shows our
predictions for $\langle L_{X} \rangle =\epsilon_X \dot{M} c^2$ as a
function of black hole mass, where $\epsilon_{X} = \alpha\epsilon_{\rm
  rad}\epsilon_{\rm bol}$, $\alpha = 0.1$ accounts for the fraction of
the inflowing matter loss to outflows from the accretion disk,
$\epsilon_{\rm rad}$ is the radiative efficiency of the accreted
matter, and $\epsilon_{\rm bol} = 0.1$ is the assumed bolometric
correction into the measured X-ray band for low luminosity AGN
(\citealt{Ho08}).  In all cases the value of $\langle L_X \rangle$ is
calculated using the accretion rate from the top line of
equation~\eqref{eq:mdot_analytic} with $\Gamma = 0.8$ for $\Mbh < 3
\times 10^7 \Msun$, and $\Gamma=-0.3 \log_{10}
\left(\frac{\Mbh}{3\times 10^7 \Msun}\right)$+0.8 for $\Mbh \geq 3
\times 10^7 \Msun$. This is functional form is designed to
approximately reproduce the behavior of $\Gamma(\Mbh)$ in the
\citet{LauerFaber+:2007a} sample.

@@

\begin{figure*}
\includegraphics[width=\textwidth]{miller.pdf}
\caption{\label{fig:miller} Average nuclear X-ray luminosity, $\langle
  L_X \rangle = \epsilon_X \Mdot c^2$, as a function of SMBH mass.
  Green lines show our prediction in the case of heating due to the
  average SFH for galaxies corresponding to each black hole mass
  (Fig.~\ref{fig:bh_growth}), calculated using $\Gamma = 0.8$ for
  $\Mbh < 3 \times 10^7 \Msun$, and $\Gamma=-0.3 \log_{10}
  \left(\frac{\Mbh}{3\times 10^7 \Msun}\right)$+0.8 for $\Mbh \geq 3
  \times 10^7 \Msun$ derived from the \citet{LauerFaber+:2007a}
  sample.  Shown for comparison are the measurements ({\it black
    stars}) and upper limits ({\it gray triangles}) for $L_X/L_{\rm
    edd}$ values, from the \citet{Miller+15} sample of early-type
  galaxies (a black line shows the best power-law fit $\langle
  L_X/L_{\rm edd}\rangle \propto M_{\bullet}^{\alpha}$, with $\alpha =
  -0.2$).  The left panel is calculated assuming a constant radiative
  efficiency $\epsilon_X=10^{-4} (\epsilon_{\rm bol}/0.1)$, while the
  right hand panel assumes $\epsilon_X= \alpha \epsilon_{\rm
    bol}\epsilon_{\rm rad}$, where $\epsilon_{\rm rad}$ is the
  radiative efficiency of low luminosity accretion disks calculated by
  \citet{Sharma+2007} (eq.~[\ref{eq:efficiency}]) and $\alpha = 0.1$
  is the fraction of the mass inflowing on large scales that reaches
  the SMBH.  Shown for comparison are the X-ray luminosities
  calculated for the maximum thermally-stable accretion rate ({\it
    dashed orange line}; eq.~[\ref{eq:Mdotmax}]); the SN Ia-regulated
  accretion rate ({\it dashed teal line}; eq.~[\ref{eq:eddr_Ia}]), and
  the Compton heating-regulated accretion rate ({\it dashed pink
    line}; eq.~[\ref{eq:MdotCa}]), again all calculated for the
  average SFH corresponding to each SMBH mass.}
\end{figure*}

If the nuclei of elliptical galaxies are heated as expected for the
average SFH of galaxies with similar mass, then to first order we
predict that $\langle L_{X}/L_{\rm edd}\rangle$ should be an 
increasing function of BH mass.  This result is in tension with the
observed down-sizing trend that the average Eddington ratio is seen to
decrease (albeit weakly) with $M_{\bullet}$, especially in the model where
$\epsilon_X$ increases with the Eddington ratio.
However, one must keep in mind the enormous uncertainty in calculating
the luminosity of the accretion flow close to the black hole in a
single waveband based on the feeding rate on larger scales.  Also, if disk
outflows indeed carry away most of the infalling mass before it reaches X-ray
producing radii (e.g.~\citealt{Blandford&Begelman99};
\citealt{Li+13}), then the angular momentum of the infalling gas might
also influence the X-ray luminosity indirectly through the inflow efficiency $\alpha$, in a way that could depend systematically on the stellar population and hence $M_{\bullet}$.  However, in general the efficiency with which inflowing matter reaches the SMBH (and hence X-ray luminosity) would naively be expected to {\it decrease} with decreasing $M_{\bullet}$ due to the increasing angular momentum of low mass galaxies, exacerbating the tension between our findings and the observed downsizing trend.

@@

\item Stellar wind heating from the average SFH may be sufficient to
  permit thermally-stable, steady-state accretion, depending on the
  accretion efficiency of the SMBH.  However, the resulting average
  Eddington-normalized accretion rate is predicted to increase with
  $M_{\bullet}$ (Fig.~\ref{fig:miller}), in tension with the (also
  weak) downsizing trend of measured X-ray luminosities in early-type
  galactic nuclei (e.g., \citealt{Miller+15}).  Thermally stable
  accretion models can reproduce the observed scatter in nuclear X-ray
  luminosities at fixed $\Mbh$ (2-4 orders of magnitude) due to the
  combination of (1) differences in the stellar wind heating rate due
  to stochastic variation in SFH between galaxies, or within one
  galaxy due to burstiness in the SFH, as in
  ~Fig.~\ref{fig:NickPlot2}; (2) Variations in the amount of inflowing
  mass which is lost to small scale outflows from the SMBH accretion
  disk, likely due to variations in the angular momentum of the
  accreting gas (cf.~\citealt{Pellegrini10}).
